High resolution, annual maps of the characteristics of smallholder-dominated croplands at national scales

### 2021-07-21 inst/manuscript/manuscript.Rmd
A few more edits.

documentclass: article
classoption: a4paper
capsize: normalsize
fontsize: 11pt
geometry: margin=1in
linenumbers: no
spacing: doublespacing
footerdate: no
note: This pre-print not yet undergone peer review. It has been submitted to the \emph{Frontiers in Artificial Intelligence}. This version will be updated as it is revised, and the final published version will be accessible through its DOI link.   
abstract: Mapping the changing characteristics of Africa's smallholder-dominated agricultural systems, including the sizes and numbers of fields, is key to understanding critical questions related to food security and other pressing socioeconomic and environmental concerns. However, accurately mapping these systems is difficult because of two key problems: 1) the spatial and temporal mismatch between most publicly satellite sensors and smallholder fields; 2) the difficulty of collecting high quality labels for training and assessing machine learning classifiers. We developed a cropland mapping approach designed to address these two problems, which we used to map Ghana’s croplands (excluding those under tree crops) for the year 2018.  To overcome the first problem, we converted daily, high resolution CubeSat (PlanetScope) imagery into two cloud-free seasonal composites covering the 2018 agricultural calendar. To address the second problem, we created a labelling platform that rigorously assesses and minimizes label error, and used it to iteratively train a machine learning classifier (Random Forests) using an active learning approach, in which we used the uncertainty in model predictions to select more informative training samples. Minimizing label errors improved model F1 scores by up to 25%, while active learning increased the F1 score by an average of 9.1% between first and last training iterations, which was 2.3% higher than models trained with randomly selected labels. We used the resulting <4 m map of cropland probabilities within a final segmentation step to delineate crop field boundaries. Using an independent map reference sample (n=1,207), we found that overall accuracies of the cropland probability and field boundary maps were 88% and 86.7%, respectively, with User’s accuracies for the cropland class of 61.2% and 78.9%, and Producer’s accuracies of 67.3% and 58.2%. Relying on the map reference sample to generate an unbiased area estimate from the field boundary map, we calculate that cropland covers 17.1 (15.4-18.9%). Using the labellers’ digitized field boundaries to estimate biases in the size and number of segments in the field boundary map, we estimate that there are 1,662,281 crop fields in Ghana with an average size of 1.73 ha. Our results demonstrate an adaptable and transferrable approach for mapping the characteristics of croplands on an annual basis and over national extents, with several features that effectively mitigate the errors inherent in remote sensing of smallholder-dominated agriculture.

@@

Beyond the imagery and computational gains, machine learning algorithms are rapidly advancing, providing large gains in classification performance [@maDeepLearningRemote2019; @MaxwellImplementationmachinelearningclassification2018]. However, the ability to take advantage of these gains is often limited by newer models' need for large training datasets, which are typically unavailable, hard to collect, or contain numerous errors [@ElmesAccountingtrainingdata2020; @maDeepLearningRemote2019; @burkeUsingSatelliteImagery2021]. To build sufficient training samples, as well as the reference data needed to objectively assess their performance (we refer collectively to both types as "labels", distinguishing between each as needed), map-makers rely heavily on visual interpretation of high resolution satellite or aerial imagery [e.g. @XiongNominal30mCropland2017; @ChenGloballandcover2015; @StehmanKeyissuesrigorous2019], as it is impractical and expensive to collect these data in the field over large areas, particularly on an ongoing basis. As such, a number of web-based platforms have been developed to collect such labels [e.g. @BeyCollectEarthLand2016; @Estesplatformcrowdsourcingcreation2016; @FritzGeoWikionlineplatform2012]. Image-drawn labels present two particular problems. The first is that they inevitably contain errors of interpretation, which can vary substantially according to the skill of the labeller, particularly over complex croplands with small field sizes [@Estesplatformcrowdsourcingcreation2016; @WaldnerConflationexpertcrowd2019]. The second problem is that visual interpretation depends on high resolution imagery (\<5 m), as fields are increasingly difficult to discern as image resolution decreases. Typically the only available source of high resolution imagery is "virtual globe" basemaps (e.g. Bing or Google Maps), which present mosaics of high resolution satellite and aerial images collected over a span of several years [@LesivCharacterizingspatialtemporal2018]. This within-mosaic temporal variation can create a temporal mismatch between the labels and the imagery being classified, which is usually from a different source (e.g. Landsat, Sentinel; @XiongNominal30mCropland2017). If a land change occurs in the interval between the two image sets (e.g. a new field was created), the label, even if accurately drawn, introduces error into the classifier. This source of error may be elevated in croplands where swidden agriculture is practiced [@vanvlietThereContinuingRationale2013], or in rapidly developing agricultural frontiers [@ZengHighlandcroplandexpansion2018].
Despite the high potential for it, label error is often not considered during model training and map accuracy assessment, resulting not only in the potential for maps to be misused or misinterpreted, but in missed opportunities to improve model performance [@ElmesAccountingtrainingdata2020; @Esteslargeareaspatiallycontinuous2018; @StehmanKeyissuesrigorous2019].

@@

## Image compositing
```{r, echo=FALSE, message=FALSE, warning=FALSE}
data("image_quality")
data("tile_key")
tile_quality <- image_quality$img %>% filter(kml_type == "F") %>% 
  group_by(name, season, worker_id) %>% 
  summarize(tile = min(tile), Score = sum(score) / 12) %>% ungroup() 
tilen <- tile_quality %>% distinct(tile) %>% count() %>% pull(n)
```
The image processing component was designed for PlanetScope Analytic surface reflectance imagery [@planetteamPlanetApplicationProgram2018], which provides three visual (red, green, blue) and near-infrared bands at 3.7 m resolution at nominal daily frequency. The images are provided as ortho-rectified and converted to surface reflectance, although there are residual errors from inter-sensor differences and the radiometric normalization process [@HouborgDailyRetrievalNDVI2018], variation in the orientation of scene footprints, as well as a high frequency of cloud cover over the study region [@WilsonRemotelySensedHighResolution2016; @royGlobalAnalysisTemporal2021] that are not fully captured by the provided cloud masks. To minimize the effect of these residual errors, we developed a procedure for creating temporal composites of the primary growing and non-growing seasons within a single 12-month period. For Ghana, we defined the primary growing season as May through September, followed by the off (or dry) season from November or December through February.
We chose these two seasons because prior work shows that the contrast between them improves cropland classifications [@Debatsgeneralizedcomputervision2016], Furthermore, capturing the seasons in this sequence during the same year helps minimize differences caused by land change. The wide time intervals we used to define each season were necessary for collecting a sufficient number of images to make high quality composites, as Ghana's cloud cover renders many scenes unusable and therefore unavailable in Planet's catalog, thus the effective return interval can be substantially longer than 24 hours during the cloudiest months [@royGlobalAnalysisTemporal2021].

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
data("image_quality")
tqual_mu <- image_quality$tile %>%
  group_by(season, tile) %>% 
  summarize(score = mean(Score)) %>% 
  ungroup()
tqual_sum <- tqual_mu %>% 
  group_by(season) %>% 
  summarize(mu = mean(score), sd = sd(score))
tqmu <- tqual_sum %>% pull(mu) %>% round(2)
names(tqmu) <- tqual_sum %>% pull(season)

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE}
data("iteration_metrics")
delta_func <- function(x, rnd = 4) round((x - lag(x)) / lag(x), rnd)

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# random versus active data
data("iteration_metrics")

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
data("label_summary")  
# toofew <- label_summary$nlabels %>% filter(n < 10) %>% pull(ID)
# tsites <- label_summary$nlabels %>% filter(!ID %in% toofew) %>% 
#   group_by(kml_type) %>% 
#   summarize(tot = sum(n), mu = round(mean(n)), med = round(median(n)), 
#             stdev = round(sd(n)), lower = min(n), upper = max(n))
# 
# # training rate
# train_rate <- label_summary$nlabels %>% 
#   tidyr::pivot_wider(names_from = kml_type, values_from = n) %>% 
#   rename_all(funs(c("ID", "tr", "tref"))) %>% 
#   mutate(trate = tref / tr) %>% filter(tr > 10)
# 
# tsitesc <- tsites %>% mutate(across(where(is.numeric), fmt))
# # nvecs <- tsites$tot[1])
# # muvecs <- fmt(tsites$mu[1])

data("label_risk")
murisk <- label_risk$stats %>% #group_by(usage) %>% 
  summarize(Risk = round(mean(Risk), 3)) #%>% 
  # tidyr::pivot_wider(names_from = usage, values_from = Risk)

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
data("consensus_high_low")
V1 <- c("Accuracy", "AUC", "F1")
delta_func <- function(x, y, rnd = 1) round((x - y) / y * 100, rnd)

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
data("label_risk_metrics")

@@

```{r, echo=FALSE}
# validation statistics
data("segment_quality_stats")

@@

```{r, echo=FALSE}
data("accuracy_reports")

@@

Using the map reference sample and each map, we made two separate estimates of the total cropland area in Ghana in 2018. The cropland extent estimated from the field boundary map was `r fmt(segas[2])` km$^2$ (with a margin of error of `r fmt(segma[1])` km$^2$), or `r segcipct[1]` (`r paste(segcipct[2:3], collapse = "-")`%) of the mapped area. The estimate from the per pixel map was `r fmt(rfas[2])` km$^2$% (margin of error = `r fmt(rfma[1])` km$^2$), or `r rfcipct[1]` (`r paste(rfcipct[2:3], collapse = "-")`%) of area.

@@

# calculate bias factors for adjustment
data("segment_quality_stats")
# segment_quality_stats %>% filter(measure == "n")
biases <- segment_quality_stats %>%
  filter(Stat == "Mean" & scale == "grid") %>% 
  pivot_wider(names_from = type, values_from = value) %>% 
  mutate(biasfactor = Validation / Segmentation)
biases <- tile_key %>% group_by(aoi) %>% 
  count() %>% ungroup() %>% rename(ntiles = n) %>% 
  mutate(aoi = as.character(aoi)) %>% 
  left_join(biases, .) %>% 
  select(-scale, -Stat)

>>>>

documentclass: article
classoption: a4paper
capsize: normalsize
fontsize: 11pt
geometry: margin=1in
linenumbers: no
spacing: doublespacing
footerdate: no
note: This pre-print not yet undergone peer review. It has been submitted to \emph{Frontiers in Artificial Intelligence}. This version will be updated as it is revised, and the final published version will be accessible through its DOI link.
abstract: Mapping the changing characteristics of Africa's smallholder-dominated agricultural systems, including the sizes and numbers of fields, is key to understanding critical questions related to food security and other pressing socioeconomic and environmental concerns. However, accurately mapping these systems is difficult because of two key problems$:$ 1) the spatial and temporal mismatch between most publicly satellite sensors and smallholder fields; 2) the difficulty of collecting high quality labels for training and assessing machine learning classifiers. We developed a cropland mapping approach designed to address these two problems, which we used to map Ghana’s croplands (excluding those under tree crops) for the year 2018. To overcome the first problem, we converted daily, high resolution CubeSat (PlanetScope) imagery into two cloud-free seasonal composites covering the 2018 agricultural calendar. To address the second problem, we created a labelling platform that rigorously assesses and minimizes label error, and used it to iteratively train a machine learning classifier (Random Forests) using an active learning approach, in which we used the uncertainty in model predictions to select more informative training samples. Minimizing label errors improved model F1 scores by up to 25%, while active learning increased the F1 score by an average of 9.1% between first and last training iterations, which was 2.3% higher than models trained with randomly selected labels. We used the resulting <4 m map of cropland probabilities within a final segmentation step to delineate crop field boundaries. Using an independent map reference sample (n=1,207), we found that overall accuracies of the cropland probability and field boundary maps were 88% and 86.7%, respectively, with User’s accuracies for the cropland class of 61.2% and 78.9%, and Producer’s accuracies of 67.3% and 58.2%. Relying on the map reference sample to generate an unbiased area estimate from the field boundary map, we calculate that cropland covers 17.1 (15.4-18.9%). Using the labellers’ digitized field boundaries to estimate biases in the size and number of segments in the field boundary map, we estimate that there are 1,662,281 crop fields in Ghana with an average size of 1.73 ha. Our results demonstrate an adaptable and transferrable approach for mapping the characteristics of croplands on an annual basis and over national extents, with several features that effectively mitigate the errors inherent in remote sensing of smallholder-dominated agriculture.

@@

Beyond the imagery and computational gains, machine learning algorithms are rapidly advancing, providing large gains in classification performance [@maDeepLearningRemote2019; @MaxwellImplementationmachinelearningclassification2018]. However, the ability to take advantage of these gains is often limited by newer models' need for large training datasets, which are typically unavailable, hard to collect, or contain numerous errors [@ElmesAccountingtrainingdata2020; @maDeepLearningRemote2019; @burkeUsingSatelliteImagery2021]. To build sufficient training samples, as well as the reference data needed to objectively assess their performance (we refer collectively to both types as "labels", distinguishing between each as needed), map-makers rely heavily on visual interpretation of high resolution satellite or aerial imagery [e.g. @XiongNominal30mCropland2017; @ChenGloballandcover2015; @StehmanKeyissuesrigorous2019], as it is impractical and expensive to collect these data in the field over large areas, particularly on an ongoing basis. Consequently, a number of web-based platforms have been developed to collect such labels [e.g. @BeyCollectEarthLand2016; @Estesplatformcrowdsourcingcreation2016; @FritzGeoWikionlineplatform2012]. Image-drawn labels present two particular problems. The first is that they inevitably contain errors of interpretation, which can vary substantially according to the skill of the labeller, particularly over complex croplands with small field sizes [@Estesplatformcrowdsourcingcreation2016; @WaldnerConflationexpertcrowd2019]. The second problem is that visual interpretation depends on high resolution imagery (\<5 m), as fields are increasingly difficult to discern as image resolution decreases. Typically the only available source of high resolution imagery is "virtual globe" basemaps (e.g. Bing or Google Maps), which present mosaics of high resolution satellite and aerial images collected over a span of several years [@LesivCharacterizingspatialtemporal2018]. This within-mosaic temporal variation can create a temporal mismatch between the labels and the imagery being classified, which is usually from a different source (e.g. Landsat, Sentinel; @XiongNominal30mCropland2017). If a land change occurs in the interval between the two image sets (e.g. a new field was created), the label, even if accurately drawn, introduces error into the classifier. This source of error may be elevated in croplands where swidden agriculture is practiced [@vanvlietThereContinuingRationale2013], or in rapidly developing agricultural frontiers [@ZengHighlandcroplandexpansion2018].
Despite the high potential for it, label error is often not considered during model training and map accuracy assessment, resulting not only in the potential for maps to be misused or misinterpreted, but in missed opportunities to improve model performance [@ElmesAccountingtrainingdata2020; @Esteslargeareaspatiallycontinuous2018; @StehmanKeyissuesrigorous2019].

@@

## Image compositing
```{r, echo=FALSE, message=FALSE, warning=FALSE}
data(image_quality)
data(tile_key)
tile_quality <- image_quality$img %>% filter(kml_type == "F") %>% 
  group_by(name, season, worker_id) %>% 
  summarize(tile = min(tile), Score = sum(score) / 12) %>% ungroup() 
tilen <- tile_quality %>% distinct(tile) %>% count() %>% pull(n)
```
The image processing component was designed for PlanetScope Analytic surface reflectance imagery [@planetteamPlanetApplicationProgram2018], which provides three visual (red, green, blue) and near-infrared bands at 3.7 m resolution at nominal daily frequency. The images are provided as ortho-rectified and converted to surface reflectance, although there are residual errors from inter-sensor differences and the radiometric normalization process [@HouborgDailyRetrievalNDVI2018], variation in the orientation of scene footprints, as well as a high frequency of cloud cover over the study region [@WilsonRemotelySensedHighResolution2016; @royGlobalAnalysisTemporal2021] that are not fully captured by the provided cloud masks. To minimize the effect of these residual errors, we developed a procedure for creating temporal composites of the primary growing and non-growing seasons within a single 12-month period. For Ghana, we defined the primary growing season as May through September, followed by the off (or dry) season from November or December through February.
We chose these two seasons because prior work shows that the contrast between them improves cropland classifications [@Debatsgeneralizedcomputervision2016], Furthermore, capturing the seasons in this sequence during the same year helps minimize differences caused by land change. The wide time intervals we used to define each season were necessary for collecting a sufficient number of images to make high quality composites, as Ghana's cloud cover renders many scenes unusable and therefore unavailable in Planet's catalog, thus the effective return interval can be substantially longer than 24 hours during the cloudiest months [@royGlobalAnalysisTemporal2021].

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
data(image_quality)
tqual_mu <- image_quality$tile %>%
  group_by(season, tile) %>% 
  summarize(score = mean(Score)) %>% 
  ungroup()
tqual_sum <- tqual_mu %>% 
  group_by(season) %>% 
  summarize(mu = mean(score), sd = sd(score))
tqmu <- tqual_sum %>% pull(mu) %>% round(2)
names(tqmu) <- tqual_sum %>% pull(season)

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE}
data(iteration_metrics)
delta_func <- function(x, rnd = 4) round((x - lag(x)) / lag(x), rnd)

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# random versus active data
data(iteration_metrics)

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
data(label_risk)
murisk <- label_risk$stats %>% #group_by(usage) %>% 
  summarize(Risk = round(mean(Risk), 3)) #%>% 
  # tidyr::pivot_wider(names_from = usage, values_from = Risk)

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
data(consensus_high_low)
V1 <- c("Accuracy", "AUC", "F1")
delta_func <- function(x, y, rnd = 1) round((x - y) / y * 100, rnd)

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
data(label_risk_metrics)

@@

```{r, echo=FALSE}
# validation statistics
data(segment_quality_stats)

@@

```{r, echo=FALSE}
data(accuracy_reports)

@@

Using the map reference sample and each map, we made two separate estimates of the total cropland area in Ghana in 2018. The cropland extent estimated from the field boundary map was `r fmt(segas[2])` km$^2$ (with a margin of error of `r fmt(segma[1])` km$^2$), or `r segcipct[1]`% (`r paste(segcipct[2:3], collapse = "-")`%) of the mapped area. The estimate from the per pixel map was `r fmt(rfas[2])` km$^2$ (margin of error = `r fmt(rfma[1])` km$^2$), or `r rfcipct[1]`% (`r paste(rfcipct[2:3], collapse = "-")`%) of area.

@@

# calculate bias factors for adjustment
data(segment_quality_stats)
# segment_quality_stats %>% filter(measure == "n")
biases <- segment_quality_stats %>%
  filter(Stat == "Mean" & scale == "grid") %>% 
  pivot_wider(names_from = type, values_from = value) %>% 
  mutate(biasfactor = Validation / Segmentation)
biases <- tile_key %>% group_by(aoi) %>% 
  count() %>% ungroup() %>% rename(ntiles = n) %>% 
  mutate(aoi = as.character(aoi)) %>% 
  left_join(biases, .) %>% 
  select(-scale, -Stat)
